---
title: "A checked historical country comparison"
format:
  html:
    toc: true
    embed-resources: true
params:
  countries: ["Ireland", "Brazil"]
  year: 2007
execute:
  echo: true
  error: false
---

## Question and Source

How do the selected countries compare in the recorded year? This is a
descriptive comparison of historical country-level data, not individual or
present-day evidence. The input is the teaching CSV supplied with Applied R;
see the [Gapminder package documentation](https://jennybc.github.io/gapminder/).

## Import, Select and Check

```{r}
source("analysis.R")
research_data <- read_research_data()
selection <- select_observations(research_data, params$countries, params$year)
stopifnot(nrow(selection) > 0L)
summary <- summarise_selection(selection)
```

The selection contains `r nrow(selection)` country-year observations for
`r params$year`, with `r summary$missing_lifeExp` missing life-expectancy values.
The country-year key is checked when the file is loaded. No source records are
edited or missing measurements replaced.

```{r}
selection[c("country", "year", "lifeExp", "pop")]
summary
```

The summary is an **unweighted mean of the selected observed country values**.
It is not a population-weighted estimate and does not describe individuals.

## A Figure with Context

```{r}
#| label: fig-comparison
#| fig-cap: "Life expectancy for the selected historical country-year observations."
#| fig-alt: "A column chart comparing the selected countries' recorded life expectancy in years. Missing values, if present, are disclosed and excluded from the chart only."
make_comparison(selection)
```

Use @fig-comparison alongside the table. Check one plotted value against its
original record, then write an interpretation that retains the year, units and
observation level. This comparison cannot establish why the values differ.

## Checks and Handover

Run `Rscript tests.R` before rendering. Record any changes made to this report,
the selected countries/year and the source input retained. Render again from a
fresh R session after changing a parameter; inspect the regenerated table,
figure and inline values together.

```{r}
sessionInfo()
```
