---
title: "Cleaning, Wrangling and Joining Data"
module-icon: "🧹"
module-number: "5"
author: "DRS Training"
metadata:
  teaching: 60
  exercises: 30
---

::: callout-outcomes
## 💡 Learning Outcomes

- Select and derive columns while retaining observation identity.
- Explain grouped summaries, missing-value choices and unmatched join records.
- Reshape a table and verify a round trip.
- Export a checked table and reuse one analysis across later outputs.
:::

::: callout-questions
## ❓ Questions

1. Which transformations change the meaning or number of records?
2. What evidence supports a missing-value or join decision?
3. How can the notebook and dashboard use the same checked calculation?
:::

## Structure & Agenda

| Block | Teaching | Activity | Output to keep |
|---|---:|---:|---|
| Inspect and select | 10 min | 5 min | A checked subset |
| Derive and summarise | 10 min | 5 min | An explicit summary |
| Join and account for records | 10 min | 5 min | A join audit |
| **Break** | — | **10 min break** | — |
| Reshape and verify | 10 min | 5 min | A checked round trip |
| File exchange | 10 min | 5 min | A checked export |
| Share the analysis | 10 min | 5 min | Reusable selection and checks |
| **Total** | **60 min** | **30 min** | **100 min including the break** |

This is **lesson 5 of eight**. Each five-minute activity includes a room response and debrief. Save work before the ten-minute break; use a practice copy and keep original inputs unchanged.

## Starting Point and Preparation

Use the [shared research project](../learners/files/research-project.zip) in a practice copy for lessons 5–8. Open that folder as the RStudio project. Complete lessons 1–4 and prepare the packages in [Setup](../learners/setup.qmd). The original Gapminder examples and calculations below are retained.

# Inspect the Data before Selecting

## Begin with a Research Question

A collaborator wants to compare historical life expectancy in two countries. Before producing a chart, identify the observation level, recorded years and units. These data concern countries over time, not individual participants or current conditions. The [Gapminder package documentation](https://jennybc.github.io/gapminder/) describes its source and coverage.

**Room check:** ask pairs to name the identifiers needed to distinguish a country observed in two different years.

## Load the Lesson Packages

Installation is covered in [Setup](../learners/setup.qmd). Load the packages for this R session:

```{r} 
library("rmarkdown")
library("gapminder")
library("tidyverse")
library("openxlsx")
library("gridExtra")
library("ggExtra")
library("knitr")
```

## Exploring the Gapminder Dataset

Let's take a look at the first few rows of the `gapminder` dataset.

```{r}
# View first six rows
head(gapminder)
str(gapminder)
range(gapminder$year)
stopifnot(!anyDuplicated(gapminder[c("country", "year")]))
```

## Data Manipulation with dplyr

The `dplyr` package provides functions for data manipulation, such as `select()`, `filter()`, `mutate()`, `arrange()`, `summarise()`, and `group_by()`.

As in lesson 4, `%>%` passes the preceding table into the next function. For example, `gapminder %>% select(country, year)` means `select(gapminder, country, year)`. Read a sequence of pipes from left to right and inspect the result after each transformation.

## Selecting Columns

Select specific columns using `select()`.

## Select Individual Columns

```{r}
# Select individual columns
sel_data <- gapminder %>% select(country, continent, pop)
head(sel_data)
```

## Filtering Rows

Use `filter()` to subset rows based on conditions.

## Filter for Belgium

```{r}
# Filter for Belgium
filt_data <- gapminder %>% filter(country == "Belgium") %>% head(8)
filt_data
```

## Filter with multiple conditions

```{r}
# Filter with multiple conditions
gapminder %>% 
  filter(country == "Oman" & year > 1980 & year <= 2000) %>% 
  head(8)
```

## Filter using %in% operator

```{r}
# Filter using %in% operator
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
res <- gapminder %>% filter(country %in% scandinavia)
head(res)
```

## Combining Select and Filter

Combine `select()` and `filter()` to extract specific data.

```{r}
uk_gdp <- gapminder %>%
  filter(country == "United Kingdom") %>%
  select(country, year, gdpPercap)
head(uk_gdp)
```

## Task 1: Check a Country–Year Selection (5 min)

::: callout-task
**Goal:** retain a specific comparison without losing row identity. **Input:** the packaged `gapminder` table. Work in pairs.

::: panel-tabset
##### Do the Task

1. **1 min:** check that 1977, Ireland and Brazil are present.
2. **2 min:** filter to those two country–year observations and select `country`, `year` and `lifeExp`.
3. **1 min:** compare the selected values with the original rows and record the units.
4. **1 min:** two pairs tell the room which identifiers and check they retained.

**Keep:** the checked subset and selection code. **If access fails:** annotate the instructor's displayed table.

##### Debrief

The output should have one row per selected country in the chosen year. Keep the country and year rather than two unlabelled values. An empty result requires a spelling or coverage check; do not silently substitute a nearby year.
:::
:::

# Derive and Summarise with Meaning

## Mutating Data

Use `mutate()` to create new columns and `transmute()` to keep only new columns.

## Using Mutate to add new columns

```{r}
# Add new columns
gapminder %>%  
  mutate(pop_million = pop / 1000000,
         countName = substring(country, 1, 3)) %>%
  head()
```

## Using Mutate to Keep Only New Columns

The multiplication below demonstrates a derived column. `lifeExp * pop` is not itself evidence of observed person-years or a standard population-health measure; interpreting such a quantity needs a justified analytical definition.

```{r}
# Keep only new columns
gapminder %>% transmute(total_years = lifeExp * pop) %>% head()
```

## Using Mutate to Classify GDP per Capita

The threshold of 10,000 below is an existing **teaching example**, not an official income classification. State its units and rationale before adapting it to a research analysis.

```{r}
# Classify GDP per capita
gapminder %>%
  mutate(GDP_Category = ifelse(gdpPercap > 10000, "High", "Low")) %>%
  select(country, year, gdpPercap, GDP_Category) %>%
  head(10)
```

## Arranging Data

Sort data using `arrange()`.

## Sort in ascending order

```{r}
# Sort in ascending order
gapminder %>% arrange(year, gdpPercap) %>% head()
```

## Sort in descending order

```{r}
# Sort in descending order
gapminder %>% arrange(year, desc(gdpPercap)) %>% head()
```

## Summarizing Data

Compute summary statistics with `summarise()`.

```{r}
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
gapminder %>% 
  filter(year == 1982 & country %in% scandinavia) %>% 
  summarise(n_obs = n(),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

## Grouping Data

Group data with `group_by()` for per-group summaries. The example calculates an unweighted mean across country observations for each year; it is not a population-weighted global life expectancy. Record the aggregation level and weighting choice.

```{r}
gapminder %>% 
  group_by(year) %>% 
  summarise(n_obs = n(),
            n_countries = n_distinct(country),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

## Task 2: Explain a Summary (5 min)

::: callout-task
**Goal:** explain the records contributing to a grouped statistic. **Input:** the year-level summary above. Swap operator and checker.

::: panel-tabset
##### Do the Task

1. **1 min:** choose one recorded year and identify its observations.
2. **2 min:** inspect the summary's row count, number of countries and mean life expectancy.
3. **1 min:** compare one statistic with a direct calculation on that year's source rows.
4. **1 min:** two pairs explain to the room whether the mean represents countries or individual people.

**Keep:** the summary and a sentence naming its aggregation and weighting. **If access fails:** inspect the instructor's summary and identify the records it combines.

##### Debrief

Grouping changes the observation level from country–year to year. The mean gives each country observation equal weight; interpreting it as the experience of an average individual would require a different, justified approach.
:::
:::

# Joining Data

Perform joins using `inner_join()`, `left_join()`, `right_join()` and `full_join()`. The employee and department tables below are **fictional teaching examples**. Check uniqueness and unmatched IDs before choosing the records to retain. See the official [join documentation](https://dplyr.tidyverse.org/reference/mutate-joins.html).

## Example data for joining

```{r}
# Create sample datasets
employees <- data.frame(
  id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  age = c(25, 30, 22, 28)
)
```

------------------------------------------------------------------------

```{r}
# Create sample datasets
departments <- data.frame(
  id = c(2, 3, 5, 6),
  department = c("HR", "Finance", "IT", "Sales")
)
```

## Inspect the Join Keys

```{r}
stopifnot(!anyDuplicated(employees$id))
stopifnot(!anyDuplicated(departments$id))
anti_join(employees, departments, by = "id")
anti_join(departments, employees, by = "id")
```

Unmatched IDs are part of the result to explain, not automatically errors to discard. Repeated keys could multiply rows; investigate their meaning before joining.

## Inner join

```{r}
# Perform inner join
employees %>% inner_join(departments, by = "id")
```

## left join

```{r}
# Perform left join
employees %>% left_join(departments, by = "id")
```

## right join

```{r}
# Perform right join
employees %>% right_join(departments, by = "id")
```

## full join

```{r}
# Perform full join
employees %>% full_join(departments, by = "id")
```

## Missing-Value Decisions with tidyr

The fictional table below demonstrates different missing-value operations using `tidyr`. Inspect the affected rows before choosing a policy. These operations change the data and answer different questions; they are not automatic cleaning steps.

## Sample data with missing values

```{r}
# Sample data with missing values
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie", NA, "Eve"),
  Age = c(25, NA, 30, 28, 22),
  Score = c(80, 90, NA, 75, 95)
)
```

## Remove rows with any NA

```{r}
# Remove rows with any NA
df_clean <- drop_na(df)
```

## Remove rows with NA in specific column

```{r}
# Remove rows with NA in specific column
df_clean_age <- drop_na(df, Age)
```

## Replace NA Values

The replacements of age with 0 and score with 50 are deliberately arbitrary syntax examples. They do not establish a valid imputation method; do not use them in research without an explicit, justified policy. Compare which values change and retain the original `df`.

```{r}
# Replace NA values
df_filled <- df %>% replace_na(list(Age = 0, Score = 50, Name = "Unknown"))
```

## Task 3: Choose a Join and Account for Records (5 min)

::: callout-task
**Goal:** make the retention decision visible. **Input:** the fictional employee and department tables. Work in pairs.

::: panel-tabset
##### Do the Task

1. **1 min:** predict which IDs a left join will keep.
2. **2 min:** compare the left and inner joins and inspect unmatched IDs in both directions.
3. **1 min:** record which rows each approach retains and why a missing department appears.
4. **1 min:** vote on a join that retains every employee; two pairs explain their choice.

**Keep:** the chosen join and record-accounting note. **If access fails:** join the displayed ID lists on paper.

##### Debrief

The left join retains employee IDs 1–4; IDs 1 and 4 have no matching department. The inner join keeps only IDs 2 and 3. A full join also retains department-only IDs 5 and 6. Choose according to the question and report unmatched records.
:::
:::

## Break (10 min)

Save the join audit before continuing.

# Reshape without Losing Observations

## Long and Wide Describe Different Arrangements

Keep the country-year key while presenting one measurement in different layouts. The two years are recorded in the data; no missing year is invented.

```{r}
long_table <- gapminder %>%
    filter(country %in% c("Ireland", "Brazil"), year %in% c(1952, 2007)) %>%
    select(country, year, lifeExp)
wide_table <- tidyr::pivot_wider(long_table, names_from = year,
                                values_from = lifeExp, names_prefix = "year_")
restored <- tidyr::pivot_longer(wide_table, cols = starts_with("year_"),
                              names_to = "year", names_prefix = "year_",
                              values_to = "lifeExp") %>%
    mutate(year = as.integer(year)) %>% arrange(country, year)
expected_long <- long_table %>% arrange(country, year)
stopifnot(isTRUE(all.equal(as.data.frame(restored), as.data.frame(expected_long))))
```

The integer conversion applies to year labels generated by the reshape, not to research measurements. A reshape can expose missing combinations; investigate the source before filling them. Different measurements and units should not be pooled into one unnamed numeric column.

## Task 4: Check a Reshape Round Trip (5 min)

::: callout-task
**Goal:** preserve four country-year measurements **Input:** the long and wide examples Work in pairs: one operates and one checks; swap roles between tasks.

::: panel-tabset
##### Do the Task

1. **1 min:** Predict the rows and columns in each arrangement.
2. **2 min:** Reshape to wide and back to long.
3. **1 min:** Compare keys, units and values after arranging by the same key.
4. **1 min:** Two pairs report what a failed comparison would require them to inspect.

**Keep:** both arrangements and their comparison. **If access fails:** annotate the instructor's displayed input and output with the same decisions and checks.

##### Debrief

Changing arrangement is not permission to aggregate, drop duplicates or impute missing combinations. Compare identity and measurements, not appearance alone.
:::
:::

# Read and Write Derived Files

Use `readr` to read and write delimited files, and `openxlsx` for Excel files. `res` is the Scandinavian-country subset created earlier. These are derived outputs; keep them separate from the original data. Rerunning an export replaces an existing file at the same path, so inspect the destination in a practice project first.

```{r}
dir.create("output", showWarnings = FALSE)
```

## Write to a file

```{r}
# Write to a file
write_delim(res, "output/gapminder2.txt", delim = "\t")
```

## Write to Excel

```{r}
# Write to Excel
write.xlsx(res, file = "output/gapm3.xlsx", sheetName = "gap", colNames = TRUE, rowNames = TRUE)
```

## Read from Excel

```{r}
# Read from Excel
xl <- read.xlsx("output/gapm3.xlsx", sheet = "gap")
```

## Check the Export against Its Input

```{r}
res_readback <- read_delim("output/gapminder2.txt", delim = "\t", show_col_types = FALSE)
stopifnot(nrow(res_readback) == nrow(res))
stopifnot(identical(as.character(res_readback$country), as.character(res$country)))
stopifnot(isTRUE(all.equal(res_readback$year, res$year)))
```

A CSV or delimited-text round trip does not retain every property of an R object, including factor levels. Check values and IDs as well as types. The Excel example writes row names explicitly; decide whether an extra row-label column suits a real handover.

## Task 5: Inspect a File Round Trip (5 min)

::: callout-task
**Goal:** verify that a derived table can be located and read. **Input:** `res` and the delimited export above. Work in pairs and swap roles.

::: panel-tabset
##### Do the Task

1. **1 min:** inspect the output destination and distinguish the source from the derived file.
2. **2 min:** write the delimited file and read it back.
3. **1 min:** compare row count, country–year IDs, types and one measurement.
4. **1 min:** two pairs give the room a check beyond file existence.

**Keep:** the file, code and read-back check. **If access fails:** annotate the instructor's displayed input and read-back table.

##### Debrief

A saved file is useful only when its location and meaning are clear. Record identity and a value comparison check the export; file existence alone cannot. Inspect type differences instead of silently treating them as identical.
:::
:::

# Share One Checked Analysis

## Keep the Calculation Separate from Its Presentation

The shared project contains `analysis.R`, a notebook, an app and known-value/failure checks. Read the helper rather than treating a downloaded file as an authority.

```{r}
source("analysis.R")
research_data <- read_research_data()
selection <- select_observations(research_data, c("Ireland", "Brazil"), 2007)
summary <- summarise_selection(selection)
stopifnot(nrow(selection) == 2L)
summary
```

The summary states selected, observed and missing counts. Its mean is unweighted across the selected observed country values; it is not a population-weighted estimate. Both later outputs call the same helper, while their inputs and presentation remain explicit.

## Task 6: Check the Shared Analysis (5 min)

::: callout-task
**Goal:** connect a result to its input and reusable function **Input:** the project and `analysis.R` Work in pairs: one operates and one checks; swap roles between tasks.

::: panel-tabset
##### Do the Task

1. **1 min:** Locate the source CSV and helper functions.
2. **2 min:** Select the two countries for 2007 and inspect the summary.
3. **1 min:** Compare one selected value with the original table; identify the missing-value policy.
4. **1 min:** Two pairs explain which code should be shared by a report and a dashboard.

**Keep:** the checked selection and documented helper contract. **If access fails:** annotate the instructor's displayed input and output with the same decisions and checks.

##### Debrief

Reuse helps keep a calculation consistent, but each interface still needs input validation, evidence checks and a statement of limitations.
:::
:::

# Optional Reference Bank

The established selection, string and worked exercise examples below remain available for follow-up. Plotting variants now live in [lesson 6](06-exploration_and_visualisation.qmd). They are outside the core 100-minute plan.


The core route above is the 100-minute session. The following retained variations support follow-up learning and instructor extensions; select them according to the task. They do not add sessions to the course.

## Selection Variants

## Select a Range of Columns

```{r}
# Select range of columns
sel_data <- gapminder %>% select(country, year:pop)
head(sel_data)
```

## Exclude columns

```{r}
# Exclude columns
sel_data <- gapminder %>% select(-pop)
head(sel_data)
```

## Exclude multiple columns

```{r}
# Exclude multiple columns
sel_data <- gapminder %>% select(-pop, -country)
head(sel_data)
```

## Exclude a vector of columns

```{r}
# Exclude a vector of columns
sel_data <- gapminder %>% select(-c(pop, country))
head(sel_data)
```

## Exclude a Complement of columns

```{r}
# Complement of columns
f1 <- gapminder %>% select(!c(continent, gdpPercap))
head(f1)
```

## Rename columns with select

```{r}
# Rename columns with select
sel_data <- gapminder %>% select(Life_Expectancy = lifeExp)
head(sel_data)
```

## Rename columns with rename()

```{r}
# Rename columns with rename()
sel_data <- gapminder %>% rename(Life_Expectancy = lifeExp)
head(sel_data)
```

## Optional Reference: Additional Selection Methods

Select columns using indices, ranges, or specific patterns.

## Select by index

```{r}
# Select by index
gapminder %>% select(c(2, 3)) %>% head()

```

## Select by index range

```{r}
# Select by index range
gapminder %>% select(2:3) %>% head()
```

## Exclude by labels

```{r}
# Select by labels
gapminder %>% select(c('country', 'continent')) %>% head()
```

## Exclude by labels

```{r}
# Exclude by labels
gapminder %>% select(-c('country', 'continent')) %>% head()
```

## Select columns beginning with 'co'

```{r}
# Select columns starting with 'co'
gapminder %>% select(starts_with('co')) %>% head()
```

## Select columns ending with 'p'

```{r}
# Select columns ending with 'p'
gapminder %>% select(ends_with('p')) %>% head()
```

## Select Columns Containing 'pop'

```{r}
# Select columns containing 'pop'
gapminder %>% select(contains('pop')) %>% head()
```

## String Manipulation with stringr

Perform string operations using `stringr`.

## Split string with stringr

```{r}
# Split string
text <- "Hello world, welcome to R programming!"
words <- str_split(text, " ")
print(words)
```

Here we split the sentence up into a list of words

## Access specific word with stringr

```{r}
# Access specific word
print(words[[1]][3])
```

We can print a specific word from the list.

## Extract substring with stringr

```{r}
sub_text <- str_sub(text, 7, 11)
print(sub_text)
```

we can extract a substring

## Replace substring with stringr

```{r}
# Replace substring
str_sub(text, 1, 5) <- "Hi"
print(text)
```

and we can replace a substring directly

# Optional Exercise Bank and Worked Answers

Choose exercises that consolidate the topic you need. Predict a result and save your attempt before inspecting the worked code. Exercises 1–5 practise selection; 7–11 practise sorting and summaries; 12–14 revisit joins, files and paths. Exercise 6 deliberately illustrates an unsupported request and should be discussed as a data-coverage check.

The worked answers below preserve the existing examples and analytical thresholds.

## Exercise 1: Select year, lifeExp, and country

```{r}
gapminder %>% select(year, lifeExp, country) %>% head()
```

## Exercise 2: Select all columns except year, lifeExp, and country

```{r}
gapminder %>% select(-year, -lifeExp, -country) %>% head()
```

## Exercise 3: Filter for year 2002

```{r}
gapminder %>% filter(year == 2002) %>% head()
```

## Exercise 4: Store and display data for Asian countries

```{r}
gapminder_asia <- gapminder %>% filter(continent == "Asia")
head(gapminder_asia)
```

## Exercise 5: Compare Ireland and Brazil in 1977

```{r}
# Ireland
gapminder %>% filter(year == 1977, country == "Ireland")

# Brazil
gapminder %>% filter(year == 1977, country == "Brazil")
```

## Exercise 6: Diagnose a Request the Dataset Cannot Answer

The inherited request asks for African countries in 2012 with `pop <= 3` and `lifeExp >= 70`. The packaged data end in 2007, and `pop` records people, so 3 is not three million. The code below preserves the original conditions and returns no observations. Explain both issues to the requester; do not silently change the year, population unit or threshold. A revised analysis requires an agreed question or another verified data source.

```{r}
df <- gapminder %>%
  filter(continent == "Africa" & year == 2012 & pop <= 3 & lifeExp >= 70) %>%
  select(country, continent)
df
```

## Exercise 7: Lowest life expectancy

```{r}
gapminder %>% arrange(lifeExp) %>% head(1)
```

## Exercise 8: Highest life expectancy

```{r}
gapminder %>% arrange(desc(lifeExp)) %>% head(1)
```

## Exercise 9: Shortest life expectancy in the Americas in 1962

```{r}
gapminder %>%
  mutate(lifeExpMonths = 12 * lifeExp) %>%
  filter(year == 1962, continent == "Americas") %>%
  arrange(lifeExpMonths) %>% head(1)
```

## Exercise 10: Summary statistics for life expectancy

```{r}
gapminder %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

## Exercise 11: Life expectancy by year

```{r}
gapminder %>%
  group_by(year) %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

## Exercise 12: Explore join statements

Already demonstrated in the "Joining Data" section above.

## Exercise 13: Read a file

Use the `output/gapm3.xlsx` file created in the file-writing example above:

```{r}
xl <- read.xlsx("output/gapm3.xlsx", sheet = "gap")
head(xl)
```

## Exercise 14: Get and split working directory path

```{r}
# Get working directory
wd <- getwd()
# Split by appropriate delimiter
path_parts <- str_split(wd, "/")[[1]]
# Fetch last element
last_element <- path_parts[length(path_parts)]
print(last_element)
```

# Further Information

::: callout-keypoints
## 📚 Key Points

- Retain the question, source, units and decisions with the result.
- Compare at least one output with an independent known value or source record.
- A successful run is one check; interpretation and limitations still need review.
:::

::: callout-hints
## 🔦 Hints

- Save before restarting R and rerun from the beginning.
- Keep original inputs separate from derived outputs.
- Record unresolved questions instead of silently changing data to obtain an answer.
:::

## Module Summary

You can account for selection, summary, join and reshape decisions and keep an analysis reusable. Next, explore and communicate its results visually.

## Additional Learning

- [tidyr pivoting](https://tidyr.tidyverse.org/articles/pivot.html): long/wide transformations and their assumptions.
- [Learner Additional Material](../learners/additional_material.qmd): practice, the shared project and official references.
- [Reference](../learners/reference.qmd): terms and checking reminders.

[Course Overview](../index.qmd) · [Previous](04-for_loops.qmd) · [Next](06-exploration_and_visualisation.qmd)
