---
title: "Data Analysis with tidyverse and ggplot2"
module-number: "4"
module-icon: "📊"
author: "DRS Training"
date: "July 7, 2025"
metadata:
  teaching: 60
  exercises: 30
webr:
  packages:
    - tidyverse
    - rmarkdown
    - knitr
    - gapminder
    - readxl
    - openxlsx
    - gridExtra
    - ggExtra
    - renv
---

::: callout-outcomes
## 💡 Learning Outcomes

- Inspect and select historical country–year observations with explicit identifiers.
- Derive and summarise columns while recording units, grouping and missing-value decisions.
- Check joins and exported files against their source records.
- Create and save a labelled ggplot2 figure with an explanation of its limits.
:::

::: callout-questions
## ❓ Questions

1. What does one observation represent and which comparisons can the data support?
2. Which selections, summaries and joins could change the meaning of a result?
3. How will another researcher check the table and figure against the input?
:::

## Structure & Agenda

| Block | Teaching | Activity | Output to keep |
|---|---:|---:|---|
| Inspect and select | 10 min | 5 min | A checked subset |
| Derive and summarise | 10 min | 5 min | A summary with its aggregation level |
| Join and handle missing values | 10 min | 5 min | A record-retention decision |
| **Break** | — | **10 min break** | — |
| Read and write derived files | 10 min | 5 min | A checked export |
| Specify a plot for the question | 10 min | 5 min | A labelled comparison |
| Check, save and communicate | 10 min | 5 min | A checked figure and interpretation |
| **Total** | **60 min** | **30 min** | **100 min including the break** |

This is **session 4 of four**, building on [Foundations](01-Rstudio_intro.qmd), [Functions](03-R_functions.qmd) and [Iteration](04-for_loops.qmd). All four sessions use the same 100-minute plan. Six five-minute room activities include their debriefs; the separate break is ten minutes. Additional selection, string and plotting examples are retained in an optional reference bank after the core session.

## Starting Point and Preparation

Open the project prepared in [Setup](../learners/setup.qmd), load the lesson packages and save the analysis in an R script. Use the packaged Gapminder dataset for the examples and keep original files separate from derived outputs. Pair work uses one operator and one checker; swap roles between activities. If software access fails, annotate the instructor's displayed tables and plots with the same selection and checks.

# Inspect the Data before Selecting

## Begin with a Research Question

A collaborator wants to compare historical life expectancy in two countries. Before producing a chart, identify the observation level, recorded years and units. These data concern countries over time, not individual participants or current conditions. The [Gapminder package documentation](https://jennybc.github.io/gapminder/) describes its source and coverage.

**Room check:** ask pairs to name the identifiers needed to distinguish a country observed in two different years.

## Load the Lesson Packages

Installation is covered in [Setup](../learners/setup.qmd). Load the packages for this R session:

```{r} 
library("rmarkdown")
library("gapminder")
library("tidyverse")
library("openxlsx")
library("gridExtra")
library("ggExtra")
library("knitr")
```

## Exploring the Gapminder Dataset

Let's take a look at the first few rows of the `gapminder` dataset.

```{r}
# View first six rows
head(gapminder)
str(gapminder)
range(gapminder$year)
stopifnot(!anyDuplicated(gapminder[c("country", "year")]))
```

## Data Manipulation with dplyr

The `dplyr` package provides functions for data manipulation, such as `select()`, `filter()`, `mutate()`, `arrange()`, `summarise()`, and `group_by()`.

As in session 3, `%>%` passes the preceding table into the next function. For example, `gapminder %>% select(country, year)` means `select(gapminder, country, year)`. Read a sequence of pipes from left to right and inspect the result after each transformation.

## Selecting Columns

Select specific columns using `select()`.

## Select Individual Columns

```{r}
# Select individual columns
sel_data <- gapminder %>% select(country, continent, pop)
head(sel_data)
```

## Filtering Rows

Use `filter()` to subset rows based on conditions.

## Filter for Belgium

```{r}
# Filter for Belgium
filt_data <- gapminder %>% filter(country == "Belgium") %>% head(8)
filt_data
```

## Filter with multiple conditions

```{r}
# Filter with multiple conditions
gapminder %>% 
  filter(country == "Oman" & year > 1980 & year <= 2000) %>% 
  head(8)
```

## Filter using %in% operator

```{r}
# Filter using %in% operator
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
res <- gapminder %>% filter(country %in% scandinavia)
head(res)
```

## Combining Select and Filter

Combine `select()` and `filter()` to extract specific data.

```{r}
uk_gdp <- gapminder %>%
  filter(country == "United Kingdom") %>%
  select(country, year, gdpPercap)
head(uk_gdp)
```

## Task 1: Check a Country–Year Selection (5 min)

::: callout-task
**Goal:** retain a specific comparison without losing row identity. **Input:** the packaged `gapminder` table. Work in pairs.

::: panel-tabset
##### Do the Task

1. **1 min:** check that 1977, Ireland and Brazil are present.
2. **2 min:** filter to those two country–year observations and select `country`, `year` and `lifeExp`.
3. **1 min:** compare the selected values with the original rows and record the units.
4. **1 min:** two pairs tell the room which identifiers and check they retained.

**Keep:** the checked subset and selection code. **If access fails:** annotate the instructor's displayed table.

##### Debrief

The output should have one row per selected country in the chosen year. Keep the country and year rather than two unlabelled values. An empty result requires a spelling or coverage check; do not silently substitute a nearby year.
:::
:::

# Derive and Summarise with Meaning

## Mutating Data

Use `mutate()` to create new columns and `transmute()` to keep only new columns.

## Using Mutate to add new columns

```{r}
# Add new columns
gapminder %>%  
  mutate(pop_million = pop / 1000000,
         countName = substring(country, 1, 3)) %>%
  head()
```

## Using Mutate to Keep Only New Columns

The multiplication below demonstrates a derived column. `lifeExp * pop` is not itself evidence of observed person-years or a standard population-health measure; interpreting such a quantity needs a justified analytical definition.

```{r}
# Keep only new columns
gapminder %>% transmute(total_years = lifeExp * pop) %>% head()
```

## Using Mutate to Classify GDP per Capita

The threshold of 10,000 below is an existing **teaching example**, not an official income classification. State its units and rationale before adapting it to a research analysis.

```{r}
# Classify GDP per capita
gapminder %>%
  mutate(GDP_Category = ifelse(gdpPercap > 10000, "High", "Low")) %>%
  select(country, year, gdpPercap, GDP_Category) %>%
  head(10)
```

## Arranging Data

Sort data using `arrange()`.

## Sort in ascending order

```{r}
# Sort in ascending order
gapminder %>% arrange(year, gdpPercap) %>% head()
```

## Sort in descending order

```{r}
# Sort in descending order
gapminder %>% arrange(year, desc(gdpPercap)) %>% head()
```

## Summarizing Data

Compute summary statistics with `summarise()`.

```{r}
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
gapminder %>% 
  filter(year == 1982 & country %in% scandinavia) %>% 
  summarise(n_obs = n(),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

## Grouping Data

Group data with `group_by()` for per-group summaries. The example calculates an unweighted mean across country observations for each year; it is not a population-weighted global life expectancy. Record the aggregation level and weighting choice.

```{r}
gapminder %>% 
  group_by(year) %>% 
  summarise(n_obs = n(),
            n_countries = n_distinct(country),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

## Task 2: Explain a Summary (5 min)

::: callout-task
**Goal:** explain the records contributing to a grouped statistic. **Input:** the year-level summary above. Swap operator and checker.

::: panel-tabset
##### Do the Task

1. **1 min:** choose one recorded year and identify its observations.
2. **2 min:** inspect the summary's row count, number of countries and mean life expectancy.
3. **1 min:** compare one statistic with a direct calculation on that year's source rows.
4. **1 min:** two pairs explain to the room whether the mean represents countries or individual people.

**Keep:** the summary and a sentence naming its aggregation and weighting. **If access fails:** inspect the instructor's summary and identify the records it combines.

##### Debrief

Grouping changes the observation level from country–year to year. The mean gives each country observation equal weight; interpreting it as the experience of an average individual would require a different, justified approach.
:::
:::

# Joining Data

Perform joins using `inner_join()`, `left_join()`, `right_join()` and `full_join()`. The employee and department tables below are **fictional teaching examples**. Check uniqueness and unmatched IDs before choosing the records to retain. See the official [join documentation](https://dplyr.tidyverse.org/reference/mutate-joins.html).

## Example data for joining

```{r}
# Create sample datasets
employees <- data.frame(
  id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  age = c(25, 30, 22, 28)
)
```

------------------------------------------------------------------------

```{r}
# Create sample datasets
departments <- data.frame(
  id = c(2, 3, 5, 6),
  department = c("HR", "Finance", "IT", "Sales")
)
```

## Inspect the Join Keys

```{r}
stopifnot(!anyDuplicated(employees$id))
stopifnot(!anyDuplicated(departments$id))
anti_join(employees, departments, by = "id")
anti_join(departments, employees, by = "id")
```

Unmatched IDs are part of the result to explain, not automatically errors to discard. Repeated keys could multiply rows; investigate their meaning before joining.

## Inner join

```{r}
# Perform inner join
employees %>% inner_join(departments, by = "id")
```

## left join

```{r}
# Perform left join
employees %>% left_join(departments, by = "id")
```

## right join

```{r}
# Perform right join
employees %>% right_join(departments, by = "id")
```

## full join

```{r}
# Perform full join
employees %>% full_join(departments, by = "id")
```

## Missing-Value Decisions with tidyr

The fictional table below demonstrates different missing-value operations using `tidyr`. Inspect the affected rows before choosing a policy. These operations change the data and answer different questions; they are not automatic cleaning steps.

## Sample data with missing values

```{r}
# Sample data with missing values
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie", NA, "Eve"),
  Age = c(25, NA, 30, 28, 22),
  Score = c(80, 90, NA, 75, 95)
)
```

## Remove rows with any NA

```{r}
# Remove rows with any NA
df_clean <- drop_na(df)
```

## Remove rows with NA in specific column

```{r}
# Remove rows with NA in specific column
df_clean_age <- drop_na(df, Age)
```

## Replace NA Values

The replacements of age with 0 and score with 50 are deliberately arbitrary syntax examples. They do not establish a valid imputation method; do not use them in research without an explicit, justified policy. Compare which values change and retain the original `df`.

```{r}
# Replace NA values
df_filled <- df %>% replace_na(list(Age = 0, Score = 50, Name = "Unknown"))
```

## Task 3: Choose a Join and Account for Records (5 min)

::: callout-task
**Goal:** make the retention decision visible. **Input:** the fictional employee and department tables. Work in pairs.

::: panel-tabset
##### Do the Task

1. **1 min:** predict which IDs a left join will keep.
2. **2 min:** compare the left and inner joins and inspect unmatched IDs in both directions.
3. **1 min:** record which rows each approach retains and why a missing department appears.
4. **1 min:** vote on a join that retains every employee; two pairs explain their choice.

**Keep:** the chosen join and record-accounting note. **If access fails:** join the displayed ID lists on paper.

##### Debrief

The left join retains employee IDs 1–4; IDs 1 and 4 have no matching department. The inner join keeps only IDs 2 and 3. A full join also retains department-only IDs 5 and 6. Choose according to the question and report unmatched records.
:::
:::

## Break (10 min)

Save the current script and note any unmatched records or unresolved missing-value decision before the break.

# Read and Write Derived Files

Use `readr` to read and write delimited files, and `openxlsx` for Excel files. `res` is the Scandinavian-country subset created earlier. These are derived outputs; keep them separate from the original data. Rerunning an export replaces an existing file at the same path, so inspect the destination in a practice project first.

```{r}
dir.create("output", showWarnings = FALSE)
```

## Write to a file

```{r}
# Write to a file
write_delim(res, "output/gapminder2.txt", delim = "\t")
```

## Write to Excel

```{r}
# Write to Excel
write.xlsx(res, file = "output/gapm3.xlsx", sheetName = "gap", colNames = TRUE, rowNames = TRUE)
```

## Read from Excel

```{r}
# Read from Excel
xl <- read.xlsx("output/gapm3.xlsx", sheet = "gap")
```

## Check the Export against Its Input

```{r}
res_readback <- read_delim("output/gapminder2.txt", delim = "\t", show_col_types = FALSE)
stopifnot(nrow(res_readback) == nrow(res))
stopifnot(identical(as.character(res_readback$country), as.character(res$country)))
stopifnot(isTRUE(all.equal(res_readback$year, res$year)))
```

A CSV or delimited-text round trip does not retain every property of an R object, including factor levels. Check values and IDs as well as types. The Excel example writes row names explicitly; decide whether an extra row-label column suits a real handover.

## Task 4: Inspect a File Round Trip (5 min)

::: callout-task
**Goal:** verify that a derived table can be located and read. **Input:** `res` and the delimited export above. Work in pairs and swap roles.

::: panel-tabset
##### Do the Task

1. **1 min:** inspect the output destination and distinguish the source from the derived file.
2. **2 min:** write the delimited file and read it back.
3. **1 min:** compare row count, country–year IDs, types and one measurement.
4. **1 min:** two pairs give the room a check beyond file existence.

**Keep:** the file, code and read-back check. **If access fails:** annotate the instructor's displayed input and read-back table.

##### Debrief

A saved file is useful only when its location and meaning are clear. Record identity and a value comparison check the export; file existence alone cannot. Inspect type differences instead of silently treating them as identical.
:::
:::

# Visualise and Communicate with ggplot2

Create visualisations using `ggplot2`. Start with the question, observation level and units. The examples below include all recorded years unless a selection is stated, so pooled country–year plots should not be described as a single-year comparison. The smooth curve describes the plotted association; it does not establish causation. Further plotting styles form an optional reference bank.

## Scatter plot with ggplot2

```{r}
#| fig-alt: "A scatter plot shows life expectancy against GDP per capita for repeated historical country-year observations in the Gapminder dataset."
# Basic scatter plot
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

## Match the Plot to a Stated Comparison

The scatter plot above combines repeated country–year observations. For the two-country comparison from Task 1, make the selected year explicit:

```{r}
#| fig-alt: "Two bars compare historical country-level life expectancy in Brazil and Ireland in 1977, with life expectancy measured in years and the Gapminder source stated."
comparison_data <- gapminder %>%
  filter(country %in% c("Ireland", "Brazil"), year == 1977) %>%
  select(country, year, lifeExp)
stopifnot(nrow(comparison_data) == 2L)

comparison_plot <- ggplot(comparison_data, aes(x = country, y = lifeExp)) +
  geom_col() +
  labs(
    title = "Life expectancy in Ireland and Brazil, 1977",
    x = "Country",
    y = "Life expectancy (years)",
    caption = "Source: Gapminder teaching dataset; historical country-level observations"
  ) +
  theme_minimal()
comparison_plot
```

The identifiers and caption explain which observations are plotted. Select a chart according to the question; the additional plotting styles below are optional references, not a checklist of required chart types.

## Task 5: Critique a Plot Specification (5 min)

::: callout-task
**Goal:** connect a figure to its observation level and units. **Input:** the pooled scatter plot and the two-country comparison above. Work in pairs.

::: panel-tabset
##### Do the Task

1. **1 min:** identify the observations contributing to each figure.
2. **2 min:** inspect the two-country plot's labels and compare its heights with `comparison_data`.
3. **1 min:** state one question the comparison supports and one it cannot answer.
4. **1 min:** two pairs explain to the room why the selected year and units matter.

**Keep:** the plot specification and limitation. **If access fails:** annotate the displayed figures and the corresponding source rows.

##### Debrief

A historical country-level comparison is descriptive. It does not establish individual outcomes, current conditions or causation. A plotted value checked against a table can still be misinterpreted if its observation level is omitted.
:::
:::

# Check, Save and Communicate the Result

## Save the Checked Plot

```{r}
ggsave("output/country_comparison.png", plot = comparison_plot, width = 7, height = 4)
```

Inspect the saved figure as well as the on-screen one. Keep its source, analysis script and a short explanation together. Save the script, restart R and run from the beginning to discover dependencies on objects created earlier in the console.

## Task 6: Make a Figure Another Researcher Can Read (5 min)

::: callout-task
**Goal:** communicate one checked historical comparison. **Input:** your two-country subset from Task 1 or the lesson's Gapminder table. Work in pairs.

::: panel-tabset
##### Do the Task

1. **1 min:** state the question, countries and year or years used.
2. **2 min:** create a suitable plot and add descriptive labels, units and a source caption.
3. **1 min:** check one plotted value against the original row and state one interpretive limitation.
4. **1 min:** two pairs show the room the check and the limitation rather than only the finished picture.

**Keep:** the saved plot, code and a short interpretation. **If access fails:** annotate an instructor plot with the required labels, source check and limitation.

##### Debrief

A readable figure identifies its observations and units. Comparing a value with the original row checks the transformation; it does not establish a causal explanation. Distinguish historical country-level patterns from present-day or individual claims.
:::
:::

# Optional Reference Bank

The core route above is the 100-minute session. The following retained variations support follow-up learning and instructor extensions; select them according to the task. They do not add sessions to the course.

## Selection Variants

## Select a Range of Columns

```{r}
# Select range of columns
sel_data <- gapminder %>% select(country, year:pop)
head(sel_data)
```

## Exclude columns

```{r}
# Exclude columns
sel_data <- gapminder %>% select(-pop)
head(sel_data)
```

## Exclude multiple columns

```{r}
# Exclude multiple columns
sel_data <- gapminder %>% select(-pop, -country)
head(sel_data)
```

## Exclude a vector of columns

```{r}
# Exclude a vector of columns
sel_data <- gapminder %>% select(-c(pop, country))
head(sel_data)
```

## Exclude a Complement of columns

```{r}
# Complement of columns
f1 <- gapminder %>% select(!c(continent, gdpPercap))
head(f1)
```

## Rename columns with select

```{r}
# Rename columns with select
sel_data <- gapminder %>% select(Life_Expectancy = lifeExp)
head(sel_data)
```

## Rename columns with rename()

```{r}
# Rename columns with rename()
sel_data <- gapminder %>% rename(Life_Expectancy = lifeExp)
head(sel_data)
```

## Optional Reference: Additional Selection Methods

Select columns using indices, ranges, or specific patterns.

## Select by index

```{r}
# Select by index
gapminder %>% select(c(2, 3)) %>% head()

```

## Select by index range

```{r}
# Select by index range
gapminder %>% select(2:3) %>% head()
```

## Exclude by labels

```{r}
# Select by labels
gapminder %>% select(c('country', 'continent')) %>% head()
```

## Exclude by labels

```{r}
# Exclude by labels
gapminder %>% select(-c('country', 'continent')) %>% head()
```

## Select columns beginning with 'co'

```{r}
# Select columns starting with 'co'
gapminder %>% select(starts_with('co')) %>% head()
```

## Select columns ending with 'p'

```{r}
# Select columns ending with 'p'
gapminder %>% select(ends_with('p')) %>% head()
```

## Select Columns Containing 'pop'

```{r}
# Select columns containing 'pop'
gapminder %>% select(contains('pop')) %>% head()
```

## String Manipulation with stringr

Perform string operations using `stringr`.

## Split string with stringr

```{r}
# Split string
text <- "Hello world, welcome to R programming!"
words <- str_split(text, " ")
print(words)
```

Here we split the sentence up into a list of words

## Access specific word with stringr

```{r}
# Access specific word
print(words[[1]][3])
```

We can print a specific word from the list.

## Extract substring with stringr

```{r}
sub_text <- str_sub(text, 7, 11)
print(sub_text)
```

we can extract a substring

## Replace substring with stringr

```{r}
# Replace substring
str_sub(text, 1, 5) <- "Hi"
print(text)
```

and we can replace a substring directly

## Additional Plotting Styles

## Bar plot with ggplot2

```{r}
# Bar plot
ggplot(data = gapminder, aes(x = continent)) +
  geom_bar()
```

## Histogram plot with ggplot2

```{r}
# Histogram
ggplot(data = gapminder, aes(x = lifeExp)) +
  geom_histogram()
```

## Box plot with ggplot2

```{r ggplot24}
# Box plot
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```

## Violin plot with ggplot2

```{r}
# Violin plot
ggplot(gapminder, aes(x = continent, y = gdpPercap)) +
  geom_violin()
```

## Scatter plot with color aesthetic with ggplot2

```{r}
# Scatter plot with color aesthetic
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color = continent)) +
  geom_point()
```

## Scatter plot with custom color

```{r}
# Scatter plot with custom color
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(color = "red")
```

## Scatter plot with smooth line

```{r}
# Scatter plot with smooth line
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5)
```

## Faceted plot in ggplot2

```{r}
# Faceted plot
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5) +
  facet_wrap(~continent)
```

## Customised Axes

The existing scale limits below remove out-of-range values before statistical layers are calculated. This can change a smooth as well as the visible points. Inspect warnings and exclusions. If you intend only to zoom the view, the [ggplot2 coordinate documentation](https://ggplot2.tidyverse.org/reference/coord_cartesian.html) explains `coord_cartesian()`; the example limits are retained here so this distinction is visible.

```{r}
# Customized axes
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5) +
  xlab("Life Expectancy") +
  ylab("GDP Per Capita") +
  xlim(0, 70) +
  ylim(0, 100000)
```

## Box plot with customized y-axis

```{r}
# Box plot with customized y-axis
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  scale_y_continuous(name = "Life Expectancy (Years)",
                     limits = c(10, 90),
                     breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90))
```

## Box plot with customized x-axis

```{r}
# Box plot with customized x-axis
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  scale_x_discrete(name = "Continent",
                   labels = c("Africa" = "AF", "Americas" = "AM",
                              "Asia" = "AS", "Europe" = "EU", "Oceania" = "OC"))
```

## Color gradient scale

```{r}
# Color gradient scale
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = pop)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")
```

## Discrete color scale

```{r}
# Discrete color scale
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_color_manual(values = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00", "#6a3d9a"))
```

## Plot with color palette

```{r}
# Plot with color palette
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(aes(fill = continent), size = 5, shape = 21) +
  geom_smooth(alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

## Box plot with theme_classic

```{r}
# Box plot with theme_classic
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  theme_classic()
```

## Box plot with jitter

```{r}
# Box plot with jitter
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.colour = "hotpink") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4)
```

## Combined violin and box plot

```{r}
# Combined violin and box plot
ggplot(gapminder) +
  geom_violin(aes(x = continent, y = gdpPercap), scale = "width") +
  geom_boxplot(aes(x = continent, y = gdpPercap), width = 0.1)
```

## Multiple plots with gridExtra

```{r}
# Multiple plots with gridExtra
p1 <- ggplot(gapminder) + geom_boxplot(aes(x = continent, y = gdpPercap, fill = continent))
p2 <- ggplot(gapminder) + geom_violin(aes(x = continent, y = gdpPercap, fill = continent))
grid.arrange(p1, p2, nrow = 1)
```

## Marginal histogram

```{r}
# Marginal histogram
p <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, fill = continent)) +
  geom_point() +
  theme(legend.position = "none")
ggMarginal(p, type = "histogram")
```

## Saving Plots

Save a plot to a file using `ggsave`.

```{r}
p1 <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(aes(fill = continent), size = 5, shape = 21) +
  geom_smooth(alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
print(getwd())
ggsave("output/myplot.png", plot = p1)
```

# Optional Exercise Bank and Worked Answers

Choose exercises that consolidate the topic you need. Predict a result and save your attempt before inspecting the worked code. Exercises 1–5 practise selection; 7–11 practise sorting and summaries; 12–14 revisit joins, files and paths. Exercise 6 deliberately illustrates an unsupported request and should be discussed as a data-coverage check.

The worked answers below preserve the existing examples and analytical thresholds.

## Exercise 1: Select year, lifeExp, and country

```{r}
gapminder %>% select(year, lifeExp, country) %>% head()
```

## Exercise 2: Select all columns except year, lifeExp, and country

```{r}
gapminder %>% select(-year, -lifeExp, -country) %>% head()
```

## Exercise 3: Filter for year 2002

```{r}
gapminder %>% filter(year == 2002) %>% head()
```

## Exercise 4: Store and display data for Asian countries

```{r}
gapminder_asia <- gapminder %>% filter(continent == "Asia")
head(gapminder_asia)
```

## Exercise 5: Compare Ireland and Brazil in 1977

```{r}
# Ireland
gapminder %>% filter(year == 1977, country == "Ireland")

# Brazil
gapminder %>% filter(year == 1977, country == "Brazil")
```

## Exercise 6: Diagnose a Request the Dataset Cannot Answer

The inherited request asks for African countries in 2012 with `pop <= 3` and `lifeExp >= 70`. The packaged data end in 2007, and `pop` records people, so 3 is not three million. The code below preserves the original conditions and returns no observations. Explain both issues to the requester; do not silently change the year, population unit or threshold. A revised analysis requires an agreed question or another verified data source.

```{r}
df <- gapminder %>%
  filter(continent == "Africa" & year == 2012 & pop <= 3 & lifeExp >= 70) %>%
  select(country, continent)
df
```

## Exercise 7: Lowest life expectancy

```{r}
gapminder %>% arrange(lifeExp) %>% head(1)
```

## Exercise 8: Highest life expectancy

```{r}
gapminder %>% arrange(desc(lifeExp)) %>% head(1)
```

## Exercise 9: Shortest life expectancy in the Americas in 1962

```{r}
gapminder %>%
  mutate(lifeExpMonths = 12 * lifeExp) %>%
  filter(year == 1962, continent == "Americas") %>%
  arrange(lifeExpMonths) %>% head(1)
```

## Exercise 10: Summary statistics for life expectancy

```{r}
gapminder %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

## Exercise 11: Life expectancy by year

```{r}
gapminder %>%
  group_by(year) %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

## Exercise 12: Explore join statements

Already demonstrated in the "Joining Data" section above.

## Exercise 13: Read a file

Use the `output/gapm3.xlsx` file created in the file-writing example above:

```{r}
xl <- read.xlsx("output/gapm3.xlsx", sheet = "gap")
head(xl)
```

## Exercise 14: Get and split working directory path

```{r}
# Get working directory
wd <- getwd()
# Split by appropriate delimiter
path_parts <- str_split(wd, "/")[[1]]
# Fetch last element
last_element <- path_parts[length(path_parts)]
print(last_element)
```

# Further Information

::: callout-keypoints
## 📚 Key Points

- Inspect identifiers, units and available years before selecting observations.
- Record grouping, weighting, unmatched joins and missing-value decisions explicitly.
- Save derived tables and figures separately and check them against source records.
:::

::: callout-hints
## 🔦 Hints

- An empty subset may expose a question the dataset cannot answer.
- Scale limits can remove observations; inspect warnings before interpreting a plot.
- Save the script, restart R and run it from the beginning to check hidden dependencies.
:::

## Module Summary

You have selected and summarised records, accounted for join behaviour and produced a checked plot. Keep the saved analysis and its assumptions together. For further practice, use the [learner Gapminder workflow](../learners/additional_material.qmd).

## Additional Learning

- [Reference](../learners/reference.qmd): syntax and terminology.
- [Discussion](../learners/discuss.qmd): explain missing values, selection and interpretation.
- [dplyr introduction](https://dplyr.tidyverse.org/articles/dplyr.html): transformation and grouping.
- [ggplot2 documentation](https://ggplot2.tidyverse.org/): labels, geoms and scales.
- [R for Data Science](https://r4ds.hadley.nz/): a structured next step after the four sessions.

For local practice, copy the relevant code into a saved R script in the course project. This page is a Quarto document; changing its extension to `.Rmd` is not required to work through the examples.

[Course Overview](../index.qmd) · [Previous: Iteration](04-for_loops.qmd) · [Further Practice](../learners/additional_material.qmd)
