---
title: "Tidyverse and ggplot2 Analysis with Gapminder Dataset"
author: "Data Analysis with R"
date: "July 7, 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This R Markdown document demonstrates data manipulation and visualization using the `tidyverse` suite of packages, particularly `dplyr`, `tidyr`, `readr`, `stringr`, and `ggplot2`. We will use the `gapminder` dataset to perform various data operations and create visualizations. The document also includes solutions to the provided exercises.

## Installing and Loading Packages
First, we install and load the necessary packages.

```{r install_packages, message=FALSE, warning=FALSE}
# Install tidyverse (includes dplyr, tidyr, readr, ggplot2, etc.)
install.packages("tidyverse", quietly = TRUE)
library(tidyverse)

# Install and load gapminder dataset
install.packages("gapminder", quietly = TRUE)
library(gapminder)

# Install and load additional packages
install.packages("openxlsx", quietly = TRUE)
library(openxlsx)
install.packages("stringr", quietly = TRUE)
library(stringr)
install.packages("gridExtra", quietly = TRUE)
library(gridExtra)
install.packages("ggExtra", quietly = TRUE)
library(ggExtra)
```

## Exploring the Gapminder Dataset
Let's take a look at the first few rows of the `gapminder` dataset.

```{r view_data}
head(gapminder) # View first six rows
```

## Data Manipulation with dplyr
The `dplyr` package provides functions for data manipulation, such as `select()`, `filter()`, `mutate()`, `arrange()`, `summarise()`, and `group_by()`.

### Selecting Columns
Select specific columns using `select()`.

```{r select_columns}
# Select individual columns
sel_data <- gapminder %>% select(country, continent, pop)
head(sel_data)

# Select range of columns
sel_data <- gapminder %>% select(country, year:pop)
head(sel_data)

# Exclude columns
sel_data <- gapminder %>% select(-pop)
head(sel_data)

# Exclude multiple columns
sel_data <- gapminder %>% select(-pop, -country)
head(sel_data)

# Exclude a vector of columns
sel_data <- gapminder %>% select(-c(pop, country))
head(sel_data)

# Complement of columns
f1 <- gapminder %>% select(!c(continent, gdpPercap))
head(f1)

# Rename columns with select
sel_data <- gapminder %>% select(Life_Expectancy = lifeExp)
head(sel_data)

# Rename columns with rename()
sel_data <- gapminder %>% rename(Life_Expectancy = lifeExp)
head(sel_data)
```

### Additional Selection Methods
Select columns using indices, ranges, or specific patterns.

```{r select_additional}
# Select by index
gapminder %>% select(c(2, 3)) %>% head()

# Select by index range
gapminder %>% select(2:3) %>% head()

# Select by labels
gapminder %>% select(c('country', 'continent')) %>% head()

# Exclude by labels
gapminder %>% select(-c('country', 'continent')) %>% head()

# Select columns starting with 'co'
gapminder %>% select(starts_with('co')) %>% head()

# Select columns ending with 'p'
gapminder %>% select(ends_with('p')) %>% head()

# Select columns containing 'pop'
gapminder %>% select(contains('pop')) %>% head()
```

### Filtering Rows
Use `filter()` to subset rows based on conditions.

```{r filter_rows}
# Filter for Belgium
filt_data <- gapminder %>% filter(country == "Belgium") %>% head(8)
filt_data

# Filter with multiple conditions
gapminder %>% 
  filter(country == "Oman" & year > 1980 & year <= 2000) %>% 
  head(8)

# Filter using %in% operator
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
res <- gapminder %>% filter(country %in% scandinavia)
head(res)
```

### Combining Select and Filter
Combine `select()` and `filter()` to extract specific data.

```{r combine_select_filter}
uk_gdp <- gapminder %>%
  filter(country == "United Kingdom") %>%
  select(country, year, gdpPercap)
head(uk_gdp)
```

### Mutating Data
Use `mutate()` to create new columns and `transmute()` to keep only new columns.

```{r mutate_data}
# Add new columns
gapminder %>%  
  mutate(pop_million = pop / 1000000,
         countName = substring(country, 1, 3)) %>%
  head()

# Keep only new columns
gapminder %>% transmute(total_years = lifeExp * pop) %>% head()

# Classify GDP per capita
gapminder %>%
  mutate(GDP_Category = ifelse(gdpPercap > 10000, "High", "Low")) %>%
  select(country, year, gdpPercap, GDP_Category) %>%
  head(10)
```

### Arranging Data
Sort data using `arrange()`.

```{r arrange_data}
# Sort in ascending order
gapminder %>% arrange(year, gdpPercap) %>% head()

# Sort in descending order
gapminder %>% arrange(year, desc(gdpPercap)) %>% head()
```

### Summarizing Data
Compute summary statistics with `summarise()`.

```{r summarise_data}
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
gapminder %>% 
  filter(year == 1982 & country %in% scandinavia) %>% 
  summarise(n_obs = n(),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

### Grouping Data
Group data with `group_by()` for per-group summaries.

```{r group_by_data}
gapminder %>% 
  group_by(year) %>% 
  summarise(n_obs = n(),
            n_countries = n_distinct(country),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

## Joining Data
Perform joins using `inner_join()`, `left_join()`, `right_join()`, and `full_join()`.

```{r join_data}
# Create sample datasets
employees <- data.frame(
  id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  age = c(25, 30, 22, 28)
)

departments <- data.frame(
  id = c(2, 3, 5, 6),
  department = c("HR", "Finance", "IT", "Sales")
)

# Perform joins
employees %>% inner_join(departments, by = "id")
employees %>% left_join(departments, by = "id")
employees %>% right_join(departments, by = "id")
employees %>% full_join(departments, by = "id")
```

## Reading and Writing Files
Use `readr` to read and write delimited files, and `openxlsx` for Excel files.

```{r read_write, eval=FALSE}
# Install readr
install.packages("readr", quietly = TRUE)
library(readr)

# Write to a file
write_delim(res, "gapminder2.txt", delim = "\t")

# Read from Excel
xl <- read.xlsx("gapminder.xlsx", sheet = "gap")

# Write to Excel
write.xlsx(res, file = "gapm3.xlsx", sheetName = "Sheet1", colNames = TRUE, rowNames = TRUE)
```

## Tidying Data with tidyr
Handle missing values using `tidyr`.

```{r tidyr}
# Sample data with missing values
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie", NA, "Eve"),
  Age = c(25, NA, 30, 28, 22),
  Score = c(80, 90, NA, 75, 95)
)

# Remove rows with any NA
df_clean <- drop_na(df)

# Remove rows with NA in specific column
df_clean_age <- drop_na(df, Age)

# Replace NA values
df_filled <- df %>% replace_na(list(Age = 0, Score = 50, Name = "Unknown"))
```

## String Manipulation with stringr
Perform string operations using `stringr`.

```{r stringr}
# Split string
text <- "Hello world, welcome to R programming!"
words <- str_split(text, " ")
print(words)

# Access specific word
print(words[[1]][3])

# Extract substring
sub_text <- str_sub(text, 7, 11)
print(sub_text)

# Replace substring
str_sub(text, 1, 5) <- "Hi"
print(text)
```

## Visualizations with ggplot2
Create visualizations using `ggplot2`.

```{r ggplot2}
# Basic scatter plot
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()

# Bar plot
ggplot(data = gapminder, aes(x = continent)) +
  geom_bar()

# Histogram
ggplot(data = gapminder, aes(x = lifeExp)) +
  geom_histogram()

# Box plot
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot()

# Violin plot
ggplot(gapminder, aes(x = continent, y = gdpPercap)) +
  geom_violin()

# Scatter plot with color aesthetic
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color = continent)) +
  geom_point()

# Scatter plot with custom color
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(color = "red")

# Scatter plot with smooth line
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5)

# Faceted plot
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5) +
  facet_wrap(~continent)

# Customized axes
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5) +
  xlab("Life Expectancy") +
  ylab("GDP Per Capita") +
  xlim(0, 70) +
  ylim(0, 100000)

# Box plot with customized y-axis
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  scale_y_continuous(name = "Life Expectancy (Years)",
                     limits = c(10, 90),
                     breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90))

# Box plot with customized x-axis
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  scale_x_discrete(name = "Continent",
                   labels = c("Africa" = "AF", "Americas" = "AM",
                              "Asia" = "AS", "Europe" = "EU", "Oceania" = "OC"))

# Color gradient scale
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = pop)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")

# Discrete color scale
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_color_manual(values = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00", "#6a3d9a"))

# Plot with color palette
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(aes(fill = continent), size = 5, shape = 21) +
  geom_smooth(alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

# Box plot with theme_classic
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  theme_classic()

# Box plot with jitter
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.colour = "hotpink") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4)

# Combined violin and box plot
ggplot(gapminder) +
  geom_violin(aes(x = continent, y = gdpPercap), scale = "width") +
  geom_boxplot(aes(x = continent, y = gdpPercap), width = 0.1)

# Multiple plots with gridExtra
p1 <- ggplot(gapminder) + geom_boxplot(aes(x = continent, y = gdpPercap, fill = continent))
p2 <- ggplot(gapminder) + geom_violin(aes(x = continent, y = gdpPercap, fill = continent))
grid.arrange(p1, p2, nrow = 1)

# Marginal histogram
p <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, fill = continent)) +
  geom_point() +
  theme(legend.position = "none")
ggMarginal(p, type = "histogram")
```

## Saving Plots
Save a plot to a file using `ggsave`.

```{r save_plot}
p1 <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(aes(fill = continent), size = 5, shape = 21) +
  geom_smooth(alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

ggsave("myplot.png", plot = p1)
```

## Exercise Solutions
Below are the solutions to the provided exercises.

### Exercise 1: Select year, lifeExp, and country
```{r exercise1}
gapminder %>% select(year, lifeExp, country) %>% head()
```

### Exercise 2: Select all columns except year, lifeExp, and country
```{r exercise2}
gapminder %>% select(-year, -lifeExp, -country) %>% head()
```

### Exercise 3: Filter for year 2002
```{r exercise3}
gapminder %>% filter(year == 2002) %>% head()
```

### Exercise 4: Store and display data for Asian countries
```{r exercise4}
gapminder_asia <- gapminder %>% filter(continent == "Asia")
head(gapminder_asia)
```

### Exercise 5: Compare Ireland and Brazil in 1977
```{r exercise5}
# Ireland
gapminder %>% filter(year == 1977, country == "Ireland")

# Brazil
gapminder %>% filter(year == 1977, country == "Brazil")
```

### Exercise 6: African countries in 2012 with specific criteria
```{r exercise6}
df <- gapminder %>%
  filter(continent == "Africa" & year == 2012 & pop <= 3 & lifeExp >= 70) %>%
  select(country, continent)
df
```

### Exercise 7: Lowest life expectancy
```{r exercise7}
gapminder %>% arrange(lifeExp) %>% head(1)
```

### Exercise 8: Highest life expectancy
```{r exercise8}
gapminder %>% arrange(desc(lifeExp)) %>% head(1)
```

### Exercise 9: Shortest life expectancy in the Americas in 1962
```{r exercise9}
gapminder %>%
  mutate(lifeExpMonths = 12 * lifeExp) %>%
  filter(year == 1962, continent == "Americas") %>%
  arrange(lifeExpMonths) %>% head(1)
```

### Exercise 10: Summary statistics for life expectancy
```{r exercise10}
gapminder %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

### Exercise 11: Life expectancy by year
```{r exercise11}
gapminder %>%
  group_by(year) %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

### Exercise 12: Explore join statements
Already demonstrated in the "Joining Data" section above.

### Exercise 13: Read a file
This requires a file like `gapminder.xlsx`. Assuming it's available:

```{r exercise13, eval=FALSE}
xl <- read.xlsx("gapminder.xlsx", sheet = "gap")
head(xl)
```

### Exercise 14: Get and split working directory path
```{r exercise14}
# Get working directory
wd <- getwd()
# Split by appropriate delimiter
path_parts <- str_split(wd, "/")[[1]]
# Fetch last element
last_element <- path_parts[length(path_parts)]
print(last_element)
```

## Conclusion
This document covers key `tidyverse` functionalities and `ggplot2` visualizations using the `gapminder` dataset. It includes data manipulation, file handling, string manipulation, and various plotting techniques, along with solutions to the exercises.

To download this R Markdown file, save it with a `.Rmd` extension and open it in RStudio or another R Markdown-compatible environment.

```