---
title: "For Loops"
author: "DRS Training"
output: html_document
---

# R For Loops

Loops are a useful part of many programming languages but are perhaps used in more limited ways in R due to the extensive use of "Vectorization".  

"Most of R's functions are vectorized, meaning that the function will operate on all elements of a vector without needing to loop through and act on each element one at a time. This makes writing code more concise, easy to read, and less error prone."

Vectorization approaches are also used extensively within dataframes.  
---

The general structure of a for loop is:

```{r view_data}
#| echo: true
sequence=c(1:10)
for (val in sequence)
{
  # do something
}
```

---

The sequence is a vector/list and val takes on each of element of the sequence in turn during the loop. In each iteration, the main code is evaluated in conjunction with that value.  

---

## Task 1

Copy the provided test code and run it

  
```{r view_data}
#| echo: true
  
  for(i in 1:10){
    cat(paste(i, '*', i, '=', i^2, '\n'))
  }
  
```

## Task 2

Try a vectorised alternative 

  
```{r view_data}
#| echo: true
  
  nums <- 1:10
  cat(paste(nums, '*', nums, '=', nums^2, collapse = "\n"))
  
```

---

You can investigate the commands/functions cat and paste further using R help e.g: <strong>?cat</strong>

---

**But some tasks do lend themselves to for loops, in particular reading things from or writing things to files**

## Task 3

Copy the provided test code, that saves some dataframes as files, and run it

 
 ```{r eval = FALSE}
 
 library(gapminder)
 library(dplyr)
 library(readr)
 
 continents <- as.character(unique(gapminder$continent))
 
 for(cont in continents){
   
   gm_cont_data <- gapminder %>%  dplyr::filter(continent == cont)
   
   write_csv(gm_cont_data, paste0('gapminder_', cont,'.csv' ))
   cat(paste0('gapminder_', cont,'.csv', '\n' ))
   
 }
 ```

## Task 4

Check the resulting files in the RStudio 'Files' tab

Can you see the difference between paste and paste0?

