---
title: "Tidyverse and ggplot2 Analysis with Gapminder Dataset"
author: "Data Analysis with R"
date: "July 7, 2025"
---


# Introduction

This R Markdown document demonstrates data manipulation and visualization using the `tidyverse` suite of packages, particularly `dplyr`, `tidyr`, `readr`, `stringr`, and `ggplot2`. We will use the `gapminder` dataset to perform various data operations and create visualizations. The document also includes solutions to the provided exercises.

# Installing and Loading Packages

First, we load the necessary packages using renv.

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))

pkgs <- c("rmarkdown","gapminder","tidyverse","readxl","openxlsx","gridExtra","ggExtra","knitr","renv")
need <- setdiff(pkgs, rownames(installed.packages()))
if (length(need)) install.packages(need, dependencies = TRUE)

invisible(lapply(pkgs, require, character.only = TRUE))
```

# Exploring the Gapminder Dataset

Let's take a look at the first few rows of the `gapminder` dataset.

```{r}
#| echo: true
# View first six rows
head(gapminder) 
```

# Data Manipulation with dplyr

The `dplyr` package provides functions for data manipulation, such as `select()`, `filter()`, `mutate()`, `arrange()`, `summarise()`, and `group_by()`.

# Selecting Columns

Select specific columns using `select()`.

## individual columns

```{r}
#| echo: true
# Select individual columns
sel_data <- gapminder %>% select(country, continent, pop)
head(sel_data)
```

## range of columns

```{r}
#| echo: true
# Select range of columns
sel_data <- gapminder %>% select(country, year:pop)
head(sel_data)
```

## Exclude columns

```{r}
#| echo: true
# Exclude columns
sel_data <- gapminder %>% select(-pop)
head(sel_data)
```

## Exclude multiple columns

```{r}
#| echo: true
# Exclude multiple columns
sel_data <- gapminder %>% select(-pop, -country)
head(sel_data)
```

## Exclude a vector of columns

```{r}
#| echo: true
# Exclude a vector of columns
sel_data <- gapminder %>% select(-c(pop, country))
head(sel_data)
```

## Exclude a Complement of columns

```{r}
#| echo: true
# Complement of columns
f1 <- gapminder %>% select(!c(continent, gdpPercap))
head(f1)
```

## Rename columns with select

```{r}
#| echo: true
# Rename columns with select
sel_data <- gapminder %>% select(Life_Expectancy = lifeExp)
head(sel_data)
```

## Rename columns with rename()

```{r}
#| echo: true
# Rename columns with rename()
sel_data <- gapminder %>% rename(Life_Expectancy = lifeExp)
head(sel_data)
```

# Additional Selection Methods

Select columns using indices, ranges, or specific patterns.

## Select by index

```{r}
#| echo: true
# Select by index
gapminder %>% select(c(2, 3)) %>% head()

```

## Select by index range

```{r}
#| echo: true
# Select by index range
gapminder %>% select(2:3) %>% head()
```

## Exclude by labels

```{r}
#| echo: true
# Select by labels
gapminder %>% select(c('country', 'continent')) %>% head()
```

## Exclude by labels

```{r}
#| echo: true
# Exclude by labels
gapminder %>% select(-c('country', 'continent')) %>% head()
```

## Select columns beginning with 'co'

```{r}
#| echo: true
# Select columns starting with 'co'
gapminder %>% select(starts_with('co')) %>% head()
```

## Select columns ending with 'p'

```{r}
#| echo: true
# Select columns ending with 'p'
gapminder %>% select(ends_with('p')) %>% head()
```

# \# Select columns containing 'pop'

```{r}
#| echo: true
# Select columns containing 'pop'
gapminder %>% select(contains('pop')) %>% head()
```

# Filtering Rows

Use `filter()` to subset rows based on conditions.

## Filter for Belgium

```{r}
#| echo: true
# Filter for Belgium
filt_data <- gapminder %>% filter(country == "Belgium") %>% head(8)
filt_data
```

## Filter with multiple conditions

```{r}
#| echo: true
# Filter with multiple conditions
gapminder %>% 
  filter(country == "Oman" & year > 1980 & year <= 2000) %>% 
  head(8)
```

## Filter using %in% operator

```{r}
#| echo: true
# Filter using %in% operator
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
res <- gapminder %>% filter(country %in% scandinavia)
head(res)
```

# Combining Select and Filter

Combine `select()` and `filter()` to extract specific data.

```{r}
#| echo: true
uk_gdp <- gapminder %>%
  filter(country == "United Kingdom") %>%
  select(country, year, gdpPercap)
head(uk_gdp)
```

# Mutating Data

Use `mutate()` to create new columns and `transmute()` to keep only new columns.

# Using Mutate to add new columns

```{r}
#| echo: true
# Add new columns
gapminder %>%  
  mutate(pop_million = pop / 1000000,
         countName = substring(country, 1, 3)) %>%
  head()
```

# Using Mutate to Keep only new columns

```{r}
#| echo: true
# Keep only new columns
gapminder %>% transmute(total_years = lifeExp * pop) %>% head()
```

# Using Mutate to Classify GDP per capita

```{r}
#| echo: true
# Classify GDP per capita
gapminder %>%
  mutate(GDP_Category = ifelse(gdpPercap > 10000, "High", "Low")) %>%
  select(country, year, gdpPercap, GDP_Category) %>%
  head(10)
```

# Arranging Data

Sort data using `arrange()`.

## Sort in ascending order

```{r}
#| echo: true
# Sort in ascending order
gapminder %>% arrange(year, gdpPercap) %>% head()
```

## Sort in descending order

```{r}
#| echo: true
# Sort in descending order
gapminder %>% arrange(year, desc(gdpPercap)) %>% head()
```

## Summarizing Data

Compute summary statistics with `summarise()`.

```{r}
#| echo: true
scandinavia <- c("Denmark", "Iceland", "Norway", "Sweden", "Finland")
gapminder %>% 
  filter(year == 1982 & country %in% scandinavia) %>% 
  summarise(n_obs = n(),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

## Grouping Data

Group data with `group_by()` for per-group summaries.

```{r}
#| echo: true
gapminder %>% 
  group_by(year) %>% 
  summarise(n_obs = n(),
            n_countries = n_distinct(country),
            total_pop = sum(pop),
            mean_life_exp = mean(lifeExp),
            range_life_exp = max(lifeExp) - min(lifeExp))
```

# Joining Data

Perform joins using `inner_join()`, `left_join()`, `right_join()`, and `full_join()`.

## Example data for joining

```{r}
#| echo: true
# Create sample datasets
employees <- data.frame(
  id = c(1, 2, 3, 4),
  name = c("Alice", "Bob", "Charlie", "David"),
  age = c(25, 30, 22, 28)
)
```

------------------------------------------------------------------------

```{r}
#| echo: true
# Create sample datasets
departments <- data.frame(
  id = c(2, 3, 5, 6),
  department = c("HR", "Finance", "IT", "Sales")
)
```

## Inner join

```{r}
#| echo: true
# Perform inner join
employees %>% inner_join(departments, by = "id")
```

## left join

```{r}
#| echo: true
# Perform left join
employees %>% left_join(departments, by = "id")
```

## right join

```{r}
#| echo: true
# Perform right join
employees %>% right_join(departments, by = "id")
```

## full join

```{r}
#| echo: true
# Perform full join
employees %>% full_join(departments, by = "id")
```

# Reading and Writing Files

Use `readr` to read and write delimited files, and `openxlsx` for Excel files.

## Write to a file

```{r}
#| echo: true
# Write to a file
write_delim(res, "data/gapminder2.txt", delim = "\t")
```

## Write to Excel

```{r}
#| echo: true
# Write to Excel
write.xlsx(res, file = "data/gapm3.xlsx", sheetName = "gap", colNames = TRUE, rowNames = TRUE)
```

## Read from Excel

```{r}
#| echo: true
# Read from Excel
xl <- read.xlsx("data/gapm3.xlsx", sheet = "gap")
```

# Tidying Data with tidyr

Handle missing values using `tidyr`.

## Sample data with missing values

```{r}
#| echo: true
# Sample data with missing values
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie", NA, "Eve"),
  Age = c(25, NA, 30, 28, 22),
  Score = c(80, 90, NA, 75, 95)
)
```

## Remove rows with any NA

```{r}
#| echo: true
# Remove rows with any NA
df_clean <- drop_na(df)
```

## Remove rows with NA in specific column

```{r}
#| echo: true
# Remove rows with NA in specific column
df_clean_age <- drop_na(df, Age)
```

## Replace NA values

```{r}
#| echo: true
# Replace NA values
df_filled <- df %>% replace_na(list(Age = 0, Score = 50, Name = "Unknown"))
```

# String Manipulation with stringr

Perform string operations using `stringr`.

## Split string with stringr

```{r}
#| echo: true
# Split string
text <- "Hello world, welcome to R programming!"
words <- str_split(text, " ")
print(words)
```

Here we split the sentence up into a list of words

## Access specific word with stringr

```{r}
#| echo: true
# Access specific word
print(words[[1]][3])
```

We can print specific works in the list

## Extract substring with stringr

`{r stringr3}#| echo: true #| echo: true # Extract substring sub_text <- str_sub(text, 7, 11) print(sub_text)`

we can extract a substring

## Replace substring with stringr

```{r}
#| echo: true
# Replace substring
str_sub(text, 1, 5) <- "Hi"
print(text)
```

and we can replace a substring directly

# Visualizations with ggplot2

Create visualizations using `ggplot2`.

## Scatter plot with ggplot2

```{r}
#| echo: true
# Basic scatter plot
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point()
```

## Bar plot with ggplot2

```{r}
#| echo: true
# Bar plot
ggplot(data = gapminder, aes(x = continent)) +
  geom_bar()
```

## Histogram plot with ggplot2

```{r}
#| echo: true
# Histogram
ggplot(data = gapminder, aes(x = lifeExp)) +
  geom_histogram()
```

## Box plot with ggplot2

```{r ggplot24}
#| echo: true
# Box plot
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot()
```

## Violin plot with ggplot2

```{r}
#| echo: true
# Violin plot
ggplot(gapminder, aes(x = continent, y = gdpPercap)) +
  geom_violin()
```

## Scatter plot with color aesthetic with ggplot2

```{r}
#| echo: true
# Scatter plot with color aesthetic
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, color = continent)) +
  geom_point()
```

## Scatter plot with custom color

```{r}
#| echo: true
# Scatter plot with custom color
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(color = "red")
```

## Scatter plot with smooth line

```{r}
#| echo: true
# Scatter plot with smooth line
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5)
```

## Faceted plot in ggplot2

```{r}
#| echo: true
# Faceted plot
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5) +
  facet_wrap(~continent)
```

## Customized axes

```{r}
#| echo: true
# Customized axes
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_smooth(alpha = 0.5) +
  xlab("Life Expectancy") +
  ylab("GDP Per Capita") +
  xlim(0, 70) +
  ylim(0, 100000)
```

## Box plot with customized y-axis

```{r}
#| echo: true
# Box plot with customized y-axis
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  scale_y_continuous(name = "Life Expectancy (Years)",
                     limits = c(10, 90),
                     breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90))
```

## Box plot with customized x-axis

```{r}
#| echo: true
# Box plot with customized x-axis
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  scale_x_discrete(name = "Continent",
                   labels = c("Africa" = "AF", "Americas" = "AM",
                              "Asia" = "AS", "Europe" = "EU", "Oceania" = "OC"))
```

## Color gradient scale

```{r}
#| echo: true
# Color gradient scale
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = pop)) +
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")
```

## Discrete color scale

```{r}
#| echo: true
# Discrete color scale
ggplot(data = gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  scale_color_manual(values = c("#1f78b4", "#33a02c", "#e31a1c", "#ff7f00", "#6a3d9a"))
```

## Plot with color palette

```{r}
#| echo: true
# Plot with color palette
ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(aes(fill = continent), size = 5, shape = 21) +
  geom_smooth(alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
```

## Box plot with theme_classic

```{r}
#| echo: true
# Box plot with theme_classic
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot() +
  theme_classic()
```

## Box plot with jitter

```{r}
#| echo: true
# Box plot with jitter
ggplot(gapminder, aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.colour = "hotpink") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4)
```

## Combined violin and box plot

```{r}
#| echo: true
# Combined violin and box plot
ggplot(gapminder) +
  geom_violin(aes(x = continent, y = gdpPercap), scale = "width") +
  geom_boxplot(aes(x = continent, y = gdpPercap), width = 0.1)
```

## Multiple plots with gridExtra

```{r}
#| echo: true
# Multiple plots with gridExtra
p1 <- ggplot(gapminder) + geom_boxplot(aes(x = continent, y = gdpPercap, fill = continent))
p2 <- ggplot(gapminder) + geom_violin(aes(x = continent, y = gdpPercap, fill = continent))
grid.arrange(p1, p2, nrow = 1)
```

## Marginal histogram

```{r}
#| echo: true
# Marginal histogram
p <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap, fill = continent)) +
  geom_point() +
  theme(legend.position = "none")
ggMarginal(p, type = "histogram")
```

# Saving Plots

Save a plot to a file using `ggsave`.

```{r}
#| echo: true
p1 <- ggplot(data = gapminder, aes(x = lifeExp, y = gdpPercap)) +
  geom_point(aes(fill = continent), size = 5, shape = 21) +
  geom_smooth(alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()
print(getwd())
ggsave("myplot.png", plot = p1)
```

# Exercise Solutions

Below are the solutions to the provided exercises.

## Exercise 1: Select year, lifeExp, and country

```{r}
#| echo: true
gapminder %>% select(year, lifeExp, country) %>% head()
```

## Exercise 2: Select all columns except year, lifeExp, and country

```{r}
#| echo: true
gapminder %>% select(-year, -lifeExp, -country) %>% head()
```

## Exercise 3: Filter for year 2002

```{r}
#| echo: true
gapminder %>% filter(year == 2002) %>% head()
```

## Exercise 4: Store and display data for Asian countries

```{r}
#| echo: true
gapminder_asia <- gapminder %>% filter(continent == "Asia")
head(gapminder_asia)
```

## Exercise 5: Compare Ireland and Brazil in 1977

```{r}
#| echo: true
# Ireland
gapminder %>% filter(year == 1977, country == "Ireland")

# Brazil
gapminder %>% filter(year == 1977, country == "Brazil")
```

## Exercise 6: African countries in 2012 with specific criteria

```{r}
#| echo: true
df <- gapminder %>%
  filter(continent == "Africa" & year == 2012 & pop <= 3 & lifeExp >= 70) %>%
  select(country, continent)
df
```

## Exercise 7: Lowest life expectancy

```{r}
#| echo: true
gapminder %>% arrange(lifeExp) %>% head(1)
```

## Exercise 8: Highest life expectancy

```{r}
#| echo: true
gapminder %>% arrange(desc(lifeExp)) %>% head(1)
```

## Exercise 9: Shortest life expectancy in the Americas in 1962

```{r}
#| echo: true
gapminder %>%
  mutate(lifeExpMonths = 12 * lifeExp) %>%
  filter(year == 1962, continent == "Americas") %>%
  arrange(lifeExpMonths) %>% head(1)
```

## Exercise 10: Summary statistics for life expectancy

```{r}
#| echo: true
gapminder %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

## Exercise 11: Life expectancy by year

```{r}
#| echo: true
gapminder %>%
  group_by(year) %>%
  summarise(
    lifeExp_min = min(lifeExp),
    lifeExp_max = max(lifeExp),
    lifeExp_mean = mean(lifeExp)
  )
```

## Exercise 12: Explore join statements

Already demonstrated in the "Joining Data" section above.

## Exercise 13: Read a file

This requires a file like `gapminder.xlsx`. Assuming it's available:

```{r}
#| echo: true
xl <- read.xlsx("data/gapm3.xlsx", sheet = "gap")
head(xl)
```

## Exercise 14: Get and split working directory path

```{r}
#| echo: true
# Get working directory
wd <- getwd()
# Split by appropriate delimiter
path_parts <- str_split(wd, "/")[[1]]
# Fetch last element
last_element <- path_parts[length(path_parts)]
print(last_element)
```

# Conclusion

This document covers key `tidyverse` functionalities and `ggplot2` visualizations using the `gapminder` dataset. It includes data manipulation, file handling, string manipulation, and various plotting techniques, along with solutions to the exercises.

To download this R Markdown file, save it with a `.Rmd` extension and open it in RStudio or another R Markdown-compatible environment.